---
title: "Outside Labels"
output: 
  html_document:
    toc: TRUE
    toc_float: TRUE
    toc_collapsed: TRUE
    toc_depth: 3
    number_sections: TRUE
    theme: lumen
---

```{r, include = FALSE}
require(mosaic)
require(Lock5Data)
data(SalaryGender)
require(gridExtra)
theme_set(theme_minimal())
require(forcats)
```

Outside labels refers to labeling all items on the "outside" of the graph. These include X and Y axis labels, titles, subtitles, captions, and legend titles. Things NOT included are labeling elements inside the graph (e.g. labeling specific points, specific lines, areas, etc.)

To adjust outside labels, we'll use the [gf_labs( )](https://www.mosaic-web.org/ggformula/reference/gf_aux.html) command.

# General Labels

```{r}
theme_set(theme_minimal())

gf_point(Sepal.Length~Sepal.Width, data = iris, color = ~Species) %>%
  gf_labs(
    y = "Sepal Length", 
    x = "Sepal Width", 
    color = "Species\nName", 
    title = "Setosa have significantly different sized sepal",
    subtitle = "While Versicolor & Virginica are more similar",
    caption = "Iris Dataset")
```

# Notes on Labels

Note that to split labels onto multiple lines, use '\\n' (example in color label above)

Other items that can be labeled are:

-   fill = "label fill legend"

-   size = "label size legend"

-   alt = "add alt text to the plot"

# Adjusting Default Labels

Sometimes the categorical labels are not what we want. For example, in the Lock5Data SalaryGender, PhD and Gender are coded as 0 and 1. This is not very clear.

Here are some ways to adjust those labels.

There are a couple problems with having the 0 and 1 labels for Gender. First, the default treating is going to be numeric - which is not ideal.

```{r}
gf_boxplot(Salary~Gender, data = SalaryGender, width = 0.4, position= position_dodge(.5)) %>%
  gf_labs(x = "Gender", title = "Gender as Numeric") 
```

At the very least, we want Gender treated as a factor. We can use the as.factor( ) command to force RStudio to read Gender as a factor. This is better, but still unclear.

```{r}
gf_boxplot(Salary~as.factor(Gender), data = SalaryGender, width = 0.4, position= position_dodge(.5)) %>%
  gf_labs(x = "Gender", title = "Gender as 0 and 1 Factor") 

```

## Adjust labels (but not data itself)

One option is to use scale_x_discrete(labels = c("0" = "F", "1" = "M")) to adjust the labels given. This will adjust what R puts on the x-axis, but it doesn't actually make Gender a factor with F and M. R Still treats it as a number. Thus, we still have to have as.factor(Gender) to get separate boxes.

```{r}
no.factor.gender = gf_boxplot(Salary~Gender, data = SalaryGender, width = 0.4, position= position_dodge(.5)) %>%
  gf_labs(x = "Gender", title = "Using scale_x_discrete Labels") %>%
  gf_refine(scale_x_discrete(labels = c("0" = "F", "1" = "M"))) 

as.factor.gender = gf_boxplot(Salary~as.factor(Gender), data = SalaryGender, width = 0.4, position= position_dodge(.5)) %>%
  gf_labs(x = "Gender", title = "Using scale_x_discrete Labels") %>%
  gf_refine(scale_x_discrete(labels = c("0" = "F", "1" = "M"))) 

grid.arrange(no.factor.gender, as.factor.gender, nrow = 1)
```

To adjust multiple labels, just adjust your scale_XXX_discrete( ) accordingly. An example below adjusts the x axis and color lables.

```{r}
gf_boxplot(Salary~as.factor(Gender), data = SalaryGender, color = ~as.factor(PhD), width = 0.4, position= position_dodge(.5)) %>%
  gf_labs(color = "PhD", x = "Gender", title = "Adjust Gender & PhD Labels") %>%
  gf_refine(
    scale_x_discrete(labels = c("0" = "Female", "1" = "Male")),
    scale_color_discrete(labels = c("0" = "No PhD", "1" = "PhD")),
    coord_cartesian(xlim = c(1,2))) 
```

## Adjust labels by adjusting the data

Our last way of adjusting the labels is to recode the variable itself. This WILL allow R to now treat Gender as a factor instead of as 0 and 1.

Note, you need to put the code to adjust your variable *before* the code for your graph as R will run these sequentially when you knit your document.

```{r}
SalaryGender$Gender = recode(SalaryGender$Gender, '0' = "Female", '1' = "Male")

gf_boxplot(Salary~Gender, data = SalaryGender, width = 0.4, position= position_dodge(.5)) %>%
  gf_labs(color = "PhD", x = "Gender", title = "Gender with Labels Adjusted") 

```

## Abbreviate labels

There are a few ways we can auto-abbreviate some labels.  [This](https://www.andrewheiss.com/blog/2022/06/23/long-labels-ggplot/) blog post has a lot of great ideas.  

**Default length (4):** scale_x_discrete(labels = abbreviate)

**Specify min length:** scale_x_discrete(labels = function(x) abbreviate(x, minlength = 1))

```{r}
gf_boxplot(Salary~Gender, data = SalaryGender, width = 0.4, position= position_dodge(.5)) %>%
  gf_labs(color = "PhD", x = "Gender", title = "Abbreviated Labels") %>%
  gf_refine(scale_x_discrete(labels = abbreviate))

gf_boxplot(Salary~Gender, data = SalaryGender, width = 0.4, position= position_dodge(.5)) %>%
  gf_labs(color = "PhD", x = "Gender", title = "Minlength Abbreviation") %>%
  gf_refine(scale_x_discrete(labels = function(x) abbreviate(x, minlength = 1)))
```

```{r}
gf_point(cty~reorder(manufacturer, displ), data = mpg) %>%
  gf_labs(x = "Manufacturer") %>%
  gf_refine(scale_x_discrete(labels = abbreviate))
```



## Dodge Labels

We can automatically dodge the labels and make them offset across multiple rows with the guide_axis(n.dodge = N) function in scale_x_*().  

```{r}
gf_point(cty~reorder(manufacturer, displ), data = mpg) %>%
  gf_labs(x = "Manufacturer") %>%
  gf_refine(scale_x_discrete(guide = guide_axis(n.dodge = 2)))
```


## Wrap Text

Another option is to automatically wrap the text (if there are spaces between words).

scale_x_discrete(labels = label_wrap(10)

# Legends

There are many things you can do with legends.  This chapter of the [ggplot2](https://ggplot2-book.org/scales-colour) book covers many more ideas than just what is listed below.

## Adjusting Legend Position

Use gf_theme(legend.position = "bottom") to adjust where the legend goes.  Choice are bottom, top, left, or right.

```{r}
gf_count(cyl~class, data = mpg, color = ~class) %>%
  gf_refine(scale_size(range = c(0.01, 12))) %>%
  gf_labs(title = "Legends at Bottom") %>%
  gf_theme(legend.position = "bottom")

gf_count(cyl~class, data = mpg, color = ~class) %>%
  gf_refine(scale_size(range = c(0.01, 12))) %>%
  gf_labs(title = "Legends at Top") %>%
  gf_theme(legend.position = "top")
```


## Removing All Legends

You can remove all legends from a single graph using the show.legend = FALSE option or using gf_theme(legend.position = "none")

```{r}
gf_count(cyl~class, data = mpg, color = ~class) %>%
  gf_refine(scale_size(range = c(0.01, 12))) %>%
  gf_labs(title = "Show All Legends") 

gf_count(cyl~class, data = mpg, color = ~class, show.legend = FALSE) %>%
  gf_refine(scale_size(range = c(0.01, 12))) %>%
  gf_labs(title = "Remove All Legends: show.legend = FALSE")

gf_count(cyl~class, data = mpg, color = ~class) %>%
  gf_refine(scale_size(range = c(0.01, 12))) %>%
  gf_labs(title = "Remove All Legends: legend.position") %>%
  gf_theme(legend.position = "none")

```

## Removing Some Legends

In order to remove just some legends, specify which legend(s) to remove using the guides(color = FALSE) command in a gf_theme( ) statement.


```{r}
gf_count(cyl~class, data = mpg, color = ~class) %>%
  gf_theme(guides(color = FALSE)) %>%
  gf_refine(scale_size(range = c(0.01, 12)))%>%
  gf_labs(title = "Remove Just Color Legend")

gf_count(cyl~class, data = mpg, color = ~class) %>%
  gf_theme(guides(size = FALSE)) %>%
  gf_refine(scale_size(range = c(0.01, 12)))%>%
  gf_labs(title = "Remove Just Size Legend")

gf_count(cyl~class, data = mpg, color = ~class) %>%
  gf_theme(guides(size = FALSE, color = FALSE)) %>%
  gf_refine(scale_size(range = c(0.01, 12)))%>%
  gf_labs(title = "Remove both in guides command")
```


