---
title: "Subsets and Zooming"
output: 
  html_document:
    toc: TRUE
    toc_float: TRUE
    toc_collapsed: TRUE
    toc_depth: 3
    number_sections: TRUE
    theme: lumen
---

```{r, include=FALSE}
require(mosaic)
```


Sometimes you want to adjust the data you see by only graphing a subset of your dataset. In order to subset your data within a graph command, you'll want to consider what type of data you have.

# Subsetting Continuous Data

**Single variable equal to a number:** subset(DatasetName, Age == 10)

**Single variable in relation to a number:** subset(DatasetName, Age \< 10)

**Single variable between 2 values:** subset(DatasetName, Age \> 3 & Age \< 10)

```{r}
theme_set(theme_minimal())

gf_point(hwy~displ, data = mpg, alpha = 0.5) %>%
  gf_labs(title = "All Data")

gf_point(hwy~displ, data = subset(mpg, year == 1999), alpha = 0.5) %>%
  gf_labs(title = "Only Year = 1999")

gf_point(hwy~displ, data = subset(mpg, year > 1999), alpha = 0.5) %>%
  gf_labs(title = "Only After 1999")

gf_point(hwy~displ, data = subset(mpg, displ <3 | displ >6), alpha = 0.5) %>%
  gf_labs(title = "Displacement on Either End (OR)")

gf_point(hwy~displ, data = subset(mpg, displ > 3 & displ < 6), alpha = 0.5) %>%
  gf_labs(title = "Displacement Between Two Values (AND)")
```

# Subsetting Cateogrical (Discrete) Data

```{r}
gf_point(hwy~displ, data = subset(mpg, class == "compact"), color = ~manufacturer, show.legend = FALSE) %>%
  gf_facet_wrap(~manufacturer) %>%
  gf_labs(title = "Facet Wrap")
```



# Multiple Subsets

AND: subset(DatasetName, year == 1999 & class == "compact")

OR: subset(DatasetName, year == 1999 | class == "compact")

List:  subset(DatasetName, manufacturer %in% c("audi", "chevrolet", "dodge"))

```{r}
gf_point(hwy~displ, data = subset(mpg, year == 1999 & class == "compact"), color = ~manufacturer, show.legend = FALSE) %>%
  gf_facet_wrap(~manufacturer) %>%
  gf_labs(title = "Facet Wrap: AND")

gf_point(hwy~displ, data = subset(mpg, year == 1999 | class == "compact"), color = ~manufacturer, show.legend = FALSE) %>%
  gf_facet_wrap(~manufacturer) %>%
  gf_labs(title = "Facet Wrap: OR")

gf_point(hwy~displ, data = subset(mpg, manufacturer %in% c("audi", "chevrolet", "dodge")), color = ~manufacturer, show.legend = FALSE) %>%
  gf_facet_wrap(~manufacturer) %>%
  gf_labs(title = "Facet Wrap: List")
```

