---
title: "All About Themes"
output: 
  html_document:
    toc: TRUE
    toc_float: TRUE
    toc_collapsed: TRUE
    toc_depth: 3
    number_sections: TRUE
    theme: lumen
---

```{r, include = FALSE}
require(mosaic)
```

Themes are going to be helpful to adjust the default. There are built-in themes already available - but you can also adjust specific elements.

# Globally set a theme

To globally set a theme, use theme_set( ) in an R Chunk. All subsequent graphs will use that theme as default.

## Global Theme Example

```{r}
theme_set(theme_bw())

gf_point(mpg~disp, data = mtcars)
```

# Manually set a theme

To manually set a theme, use the gf_theme( ) layer on any plot.

## Manual theme example

```{r}
gf_point(mpg~disp, data = mtcars) %>%
  gf_theme(theme_minimal())
```

# All ggplot themes

Here is a [list](https://ggplot2.tidyverse.org/reference/ggtheme.html) of all themes available and a preview of what they look like.

```{r}
require(gridExtra)

gray = gf_point(mpg~disp, data = mtcars) %>%
  gf_theme(theme_gray()) %>%
  gf_labs(title = "theme_gray")

bw = gf_point(mpg~disp, data = mtcars) %>%
  gf_theme(theme_bw()) %>%
  gf_labs(title = "theme_bw")

linedraw = gf_point(mpg~disp, data = mtcars) %>%
  gf_theme(theme_linedraw()) %>%
  gf_labs(title = "theme_linedraw")

light = gf_point(mpg~disp, data = mtcars) %>%
  gf_theme(theme_light()) %>%
  gf_labs(title = "theme_light")

dark = gf_point(mpg~disp, data = mtcars) %>%
  gf_theme(theme_dark()) %>%
  gf_labs(title = "theme_dark")

minimal = gf_point(mpg~disp, data = mtcars) %>%
  gf_theme(theme_minimal()) %>%
  gf_labs(title = "theme_minimal")

classic = gf_point(mpg~disp, data = mtcars) %>%
  gf_theme(theme_classic()) %>%
  gf_labs(title = "theme_classic")

void = gf_point(mpg~disp, data = mtcars) %>%
  gf_theme(theme_void()) %>%
  gf_labs(title = "theme_void")

test = gf_point(mpg~disp, data = mtcars) %>%
  gf_theme(theme_test()) %>%
  gf_labs(title = "theme_test")


grid.arrange(gray, bw, linedraw, light, dark, minimal, classic, void, test, nrow = 3 )
```

